---
layout: lesson
title: "Session 14"
output: markdown_document
---

## Topics
*


```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)

opts_chunk$set("echo" = TRUE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("message" = FALSE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

opts_chunk$set("fig.retina" = 4)
opts_chunk$set("results" = "hold")
opts_chunk$set("fig.show" = "hold")
opts_chunk$set("fig.width" = 7)
```

In the last several sessions, we looked at gender and racial/ethnic representation at pretty granular levels - the level of institution type and particular institutions. I'd like to get a bit more refined and look within a specific set of disciplines. I'm going to start the analysis with "Microbiology". To do this analysis, I need to develop a sense of how the data are reported. A "microbiologist" is a very diverse concept! They can study organisms in any of the three domains of life, viruses, interactions with animals or plants or the natural environment. They can be engineered, host-associated, or natural ecosystems. They can be studied at the gene, protein, organismal, or community level. Depending on the campus, the people that study microbes at these various levels can be one person in a Biology department or 20 in a Microbiology department or 200 across 20 different departments. How we organize data into a framework is critical to the interpretation of our data. We could be the best R programmer in the world, but if we can't get the definitions right, we're toast.

Inspecting the CIP codes in `ipeds/c2018_a.xlsx`, I see that there's some organization to them - the first two numbers correspond to a broad discipline (e.g. 26 - Biology or 27 - Math). Naturally, I immediately noticed that "26.1102 - Biostatistics" is listed under Biology and not under Math and "27.0306 - Mathematical Biology" is listed under Math rather than Biology. For the microbiology-affiliated fields I listed above, I notice that their CIP codes all start with "26.05". Among those CIP codes, there are

* 26.0502	Microbiology, General
* 26.0503	Medical Microbiology and Bacteriology
* 26.0504	Virology
* 26.0505	Parasitology
* 26.0507	Immunology
* 26.0508	Microbiology and Immunology
* 26.0599	Microbiological Sciences and Immunology, Other

Of course, there are Immunologists that don't think much about microbes and there are card carrying microbiologists in general Biology departments and *even* Engineering. This overlap between subdisciplines is why it is a bad idea to aggregate all institutions blindly. A better practice would be to pick peer institutions where one has a better understanding of where a microbiologist would most likely be trained within that institution. In other words, take this analysis with a grain of salt. We're here to learn R, not make grand statements about bias in science.

This discussion gets us pretty close to figuring out the categories we're interested in, but let's step back and see if we can do a more systematic review of where microbiology shows up in the CIP codes.

```{r}
library(tidyverse)
library(readxl)

cip_codes <- read_excel("ipeds/c2018_a.xlsx", sheet="Frequencies") %>%
	filter(varname == "CIPCODE" & codevalue != "99") %>%
	select(codevalue, valuelabel) %>%
	rename(cip_code = codevalue,
		description = valuelabel)
```

As I pointed out above, it seems pretty clear that microbiology disciplines are "supposed" to have a CIP code that starts "26.05". I'm curious whether "Microbiology" shows up in any other titles. To figure this out, we need to learn a new function - `str_detect`. This function tells us whether a character variable contains a string we are interested in. For example, consider this character vector

```{r}
baseball_teams <- c("cubs", "white sox", "red sox", "cardinals", "yankees")
```

I want to know which values in `baseball_teams` has the word "sox". I can use `str_detect`

```{r}
str_detect(baseball_teams, "sox")
```

If I wanted to know which elements in `baseball_teams` had "sox", I could use `str_which`

```{r}
str_which(baseball_teams, "sox")
```

And if I wanted to get the values that contained "sox", I would use `str_subset`....

```{r}
str_subset(baseball_teams, "sox")
```

The nice thing about `str_detect` is that because it returns a logical value, I can use it with `filter` to get those rows from a data frame that match the search string

```{r}
cip_codes %>%
	filter(str_detect(cip_code, "26.05"))
```

Let's look for microbiology elsewhere by using `str_detect` on the `description` column

```{r}
cip_codes %>%
	filter(str_detect(description, "Microb"))
```

Aha, we found a "Veterinary Microbiology and Immunology".

Let's look a bit broader for those descriptions that have "biology" in them

```{r}
cip_codes %>%
	filter(str_detect(description, "biology"))
```

of course, we could have also searched for

```{r}
cip_codes %>%
	filter(str_detect(description, "Biology"))
```

or

```{r}
cip_codes %>%
	filter(str_detect(description, "Biolog"))
```

Developing a meaningful search term is a challenge. If we are too broad (e.g. "Micro") we could get something that isn't relevant like "Data Entry/Microcomputer Applications". If we're too specific (e.g. "Biological" vs. "Biology" vs "biology"), then we risk losing out on relevant matches. To make our search insensitive to the case of the search string

```{r}
cip_codes %>%
	filter(str_detect(description, "(?i)Biolog"))
```

Notice that this will match both "Biology/Biological Sciences, General" and "Microbiology, General". The "(?i)Biolog" is what is called a "regular expression". Believe it or not, this is a relatively simple regular expression. We can use far more complicated patterns to identify dates, email addresses, genes, and more.

If it wasn't already clear, defining "biology" or "microbiology" is pretty hard! Let's stick with the "26.05" CIP code stub to define microbiology

```{r}
microbiology_cips <- read_excel("ipeds/c2018_a.xlsx", sheet="Frequencies") %>%
	filter(varname == "CIPCODE" & codevalue != "99") %>%
	select(codevalue, valuelabel) %>%
	rename(cip_code = codevalue,
		description = valuelabel) %>%
	filter(str_detect(cip_code, "26.05"))
```

Which institutions have enough microbiology on campus that they feel comfortable using one of these codes? To answer this question, I could now do an inner join between `microbiology` and the data in `ipeds/c2018_a.csv` to find the `UNITID` values that reported graduates in one of these CIP codes. Then I could join those results with the broader institutional data in `ipeds/hd2018.csv`. All of this should sound familiar by now.

```{r}
microbiology_degrees <- read_csv("ipeds/c2018_a.csv") %>%
	inner_join(microbiology_cips, ., by=c("cip_code" = "CIPCODE")) %>%
	filter(MAJORNUM == 1 & CTOTALT > 0 & (AWLEVEL == "05" | AWLEVEL == "17"))
```

Before we go all the way, let's see how many institutions claimed each type of microbiology for each degree type.

```{r}
microbiology_degrees %>%
	count(description, AWLEVEL)
```

That's a little hard for me to read, what if we make the column names the different degree types? Remember how to do that?

```{r}
microbiology_degrees %>%
	count(description, AWLEVEL) %>%
	pivot_wider(names_from=AWLEVEL, values_from=n, values_fill=list(n=0)) %>% # if we leave out `values_fill we get NA values`
	select(description, `05`, `17`)
```

From this, it looks like most bachelor's degrees in Microbiology are in "Microbiology, General" and occasionally in "Medical Microbiology and Bacteriology". On the other hand, there is more specialization at the doctorate level. Back to our question of which schools are these... I will modify `microbiology_degrees` to get the total number of graduates with a bachelor's and doctorate degree from each major as separate columns

```{r}
microbiology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALT) %>%
	pivot_wider(names_from=AWLEVEL, values_from=CTOTALT, values_fill=list(n=0))
```

This code chunk shows there are 196 institution/department combinations that awarded a microbiology degree in 2018. Are there institutions that show up multiple times?

```{r}
microbiology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALT) %>%
	pivot_wider(names_from=AWLEVEL, values_from=CTOTALT, values_fill=list(n=0)) %>%
	count(UNITID) %>%
	count(n)
```

It appears that 139 schools offer one microbiology major, 22 offer two, 3 offer three, and 1 offers four. How many does the University of Michigan offer?

```{r}
microbiology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALT) %>%
	pivot_wider(names_from=AWLEVEL, values_from=CTOTALT, values_fill=list(n=0)) %>%
	count(UNITID) %>%
	filter(UNITID == 170976)
```

3!?! I know there's "Microbiology and Immunology", "Immunology" and... I'm not sure. Let's find out.

```{r}
microbiology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALT) %>%
	pivot_wider(names_from=AWLEVEL, values_from=CTOTALT, values_fill=list(n=0)) %>%
	filter(UNITID == 170976)
```

Ah, yes. The undergrads - "Microbiology, General". Instead of looking up `UNITID` codes manually, let's bring those in with a join

```{r}
micro_degree_counts <- microbiology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALT) %>%
	pivot_wider(names_from=AWLEVEL, values_from=CTOTALT, values_fill=list(n=0)) %>%
	inner_join(read_csv("ipeds/hd2018.csv"), ., by="UNITID") %>%
	select(UNITID, INSTNM, description, `17`, `05`)
```

Now it's easier to see how many different microbiology majors these schools offer

```{r}
micro_degree_counts %>%
	count(INSTNM) %>%
	arrange(desc(n))
```

Looking through these institutions, it's pretty clear that to have a dedicated microbiology program of any type suggests that the institution is research focused. We can also see that the program to graduate the most doctorates, University of Wisconsin-Madison, only awarded 19 PhDs in 2018. Based on this, it probably doesn't make much sense to compare the representation at individual institutions for a single year. Now that we have a better handle on the data, let's see what the gender representation looks like for microbiologists awarded bachelor's and doctorate degrees in 2018.

```{r}
microbiology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALW, CTOTALT) %>%
	group_by(AWLEVEL) %>%
	summarize(f_women = sum(CTOTALW)/sum(CTOTALT))
```

The fraction of women receiving a bachelors degree and a PhD are fairly similar in microbiology. What does this fraction look like for all of the biology CIP codes? To answer this, I can modify the code chunk from above where we defined `microbiology_cips`

```{r}
biology_cips <- read_excel("ipeds/c2018_a.xlsx", sheet="Frequencies") %>%
	filter(varname == "CIPCODE" & codevalue != "99") %>%
	select(codevalue, valuelabel) %>%
	rename(cip_code = codevalue,
		description = valuelabel) %>%
	filter(str_detect(cip_code, "26"))
```

If you look at `biology_cips` you'll notice a small problem. We identified any CIP code that contained "26", which got us things like "Italian Studies" and "German Language Teacher Education". Oops. We what CIP codes that *start* with "26". We can modify our search term to use an "anchor". The `^` tells `str_detect` to look at the beginning of a string and `$` tells it to look at the end of the string.

```{r}
biology_cips <- read_excel("ipeds/c2018_a.xlsx", sheet="Frequencies") %>%
	filter(varname == "CIPCODE" & codevalue != "99") %>%
	select(codevalue, valuelabel) %>%
	rename(cip_code = codevalue,
		description = valuelabel) %>%
	filter(str_detect(cip_code, "^26"))
```

That's more like it. We can create `biology_degrees` similar to how we generated `microbiology_degrees`

```{r}
biology_degrees <- read_csv("ipeds/c2018_a.csv") %>%
	inner_join(biology_cips, ., by=c("cip_code" = "CIPCODE")) %>%
	filter(MAJORNUM == 1 & CTOTALT > 0 & (AWLEVEL == "05" | AWLEVEL == "17"))
```

And we can calculate the fraction of women across all biology for both degree levels

```{r}
biology_degrees %>%
	select(UNITID, description, AWLEVEL, CTOTALW, CTOTALT) %>%
	group_by(AWLEVEL) %>%
	summarize(f_women = sum(CTOTALW)/sum(CTOTALT))
```

Interesting, it appears that in some fields there is a larger drop in representation of women. The individual number of graduates in each sub discipline are too small to tell us much. However, if we aggregated the data at the "hundredths" place in the CIP code we could look at a broader level that is still more resolved than all of biology.

How do we extract "26.05" from "26.0502"? How do we do extract "26.01" from "26.0101" with the same command? We need another function that works with strings - `str_replace`. This function takes a set of strings, a pattern to match, and a replacement value. Here's the brute force way to do it for one CIP code

```{r}
cip_code <- "26.0502"
str_replace(cip_code, pattern="26.0502", replacement="26.05")
```

The challenge is that we can't write hundreds of `str_replace` functions. Also, if we reanalyze the data when the 2019 data are released, we may need to add more `str_replace` function calls. No fun. We need to learn a little more about regular expressions. We can use "metacharacters" to represent a generic letter, number, or space. Here's a quick set of examples...

```{r}
x <- "My phone number is (555)678-1234"
#str_replace only replaces the first matched value
str_replace_all(x, "\\w", "x") # match any alphanumeric character
```

```{r}
str_replace_all(x, "\\d", "#") # match any number
```

```{r}
str_replace_all(x, "\\s", "_") # match any space
```

```{r}
str_replace_all(x, "[()-]", "_") # match any (, ), or -
```

We can match more than a single character

```{r}
str_replace_all(x, "\\d\\d\\d\\d", "####") # match any number
```

What does that mean for our CIP code?

```{r}
str_replace(cip_code, "\\d\\d$", "")
```

Let's apply this to group by the sub fields within biology and calculate the percent drop going from bachelor's to doctorate degrees

```{r}
biology_degrees %>%
	select(UNITID, cip_code, description, AWLEVEL, CTOTALW, CTOTALT) %>%
	mutate(sub_category = str_replace(cip_code, "\\d\\d$", "")) %>%
	group_by(sub_category, AWLEVEL) %>%
	summarize(f_women = sum(CTOTALW)/sum(CTOTALT)) %>%
	ungroup() %>%
	pivot_wider(names_from=AWLEVEL, values_from=f_women) %>%
	mutate(percent_drop = 100*(`05`-`17`)/`05`) %>%
	arrange(percent_drop)
```

It appears that of all the biology disciplines, microbiology does the best job of retaining its gender balance between the bachelor's and doctorate degree awardees.


## Exercises

1\. It is interesting that after microbiology, sub field "26.13" is next in the ranking for retaining women and has more than 60% of its degrees going to women at both levels. What subfields are in that grouping? The worst at retention is "26.07". What field is that?

<input type="button" class="hideshow">
<div markdown="1" style="display:none;">
```{r}
biology_cips %>%
	filter(str_detect(cip_code, "^26.13"))
```

Ecology and evolutionary biology.

```{r}
biology_cips %>%
	filter(str_detect(cip_code, "^26.07"))
```

Animal biology
</div>

2\. Generate a slopegraph of the biology CIP codes with each line representing a different sub field (i.e. to the level of XX.XX). Can you highlight the microbiology line?

<input type="button" class="hideshow">
<div markdown="1" style="display:none;">
```{r}
biology_degrees %>%
	select(UNITID, cip_code, description, AWLEVEL, CTOTALW, CTOTALT) %>%
	mutate(sub_category = str_replace(cip_code, "\\d\\d$", "")) %>%
	group_by(sub_category, AWLEVEL) %>%
	summarize(f_women = sum(CTOTALW)/sum(CTOTALT)) %>%
	ungroup() %>%
	mutate(highlight = sub_category == "26.05",
		sub_category = fct_reorder(sub_category, highlight)) %>%
	ggplot(aes(x=AWLEVEL, y=f_women, group=sub_category, color=highlight)) +
		geom_line(show.legend=FALSE) +
		scale_color_manual(name="Highlighted",
			breaks=c(FALSE, TRUE),
			labels=c("Other Institutions", "University of Michigan-Ann Arbor"),
			values=c("gray", "blue")) +
		theme_classic()
```
</div>


3\. Can you repeat the analysis resulting in the slope graph in the last exercise, but looking at the representation of "Black and African American" awardees as a ratio of "White" awardees?

<input type="button" class="hideshow">
<div markdown="1" style="display:none;">
```{r}
biology_degrees %>%
	select(UNITID, cip_code, description, AWLEVEL, CBKAAT, CWHITT) %>%
	mutate(sub_category = str_replace(cip_code, "\\d\\d$", "")) %>%
	group_by(sub_category, AWLEVEL) %>%
	summarize(r_bkaa = sum(CBKAAT)/sum(CWHITT)) %>%
	ungroup() %>%
	mutate(highlight = sub_category == "26.05",
		sub_category = fct_reorder(sub_category, highlight)) %>%
	ggplot(aes(x=AWLEVEL, y=r_bkaa, group=sub_category, color=highlight)) +
		geom_line(show.legend=FALSE) +
		scale_color_manual(name="Highlighted",
			breaks=c(FALSE, TRUE),
			labels=c("Other Institutions", "University of Michigan-Ann Arbor"),
			values=c("gray", "blue")) +
		theme_classic()
```
</div>


4\. Of the engineering sub fields with more than 100 bachelor's and 100 doctorate degrees, which has the best retention and representation of women? The worst?

<input type="button" class="hideshow">
<div markdown="1" style="display:none;">
```{r}
engineering_cips <- read_excel("ipeds/c2018_a.xlsx", sheet="Frequencies") %>%
	filter(varname == "CIPCODE" & codevalue != "99") %>%
	select(codevalue, valuelabel) %>%
	rename(cip_code = codevalue,
		description = valuelabel) %>%
	filter(str_detect(cip_code, "^14"))

engineering_degrees <- read_csv("ipeds/c2018_a.csv") %>%
	inner_join(engineering_cips, ., by=c("cip_code" = "CIPCODE")) %>%
	filter(MAJORNUM == 1 & CTOTALT > 0 & (AWLEVEL == "05" | AWLEVEL == "17"))

engineering_degrees %>%
	select(UNITID, cip_code, description, AWLEVEL, CTOTALW, CTOTALT) %>%
	mutate(sub_category = str_replace(cip_code, "\\d\\d$", "")) %>%
	group_by(sub_category, AWLEVEL) %>%
	summarize(n_degrees = sum(CTOTALT), f_women = sum(CTOTALW)/sum(CTOTALT)) %>%
	ungroup() %>%
	pivot_wider(names_from=AWLEVEL, values_from=c(f_women, n_degrees)) %>%
	mutate(percent_drop = 100*(f_women_05-f_women_17)/f_women_05) %>%
	arrange(percent_drop) %>%
	filter(n_degrees_05 > 100 & n_degrees_17 > 100) %>%
	print(n=100)
```

Best retention...
```{r}
engineering_cips %>% filter(str_detect(cip_code, "^14.09"))
```

Computer science.

Worst retention....
```{r}
engineering_cips %>% filter(str_detect(cip_code, "^14.35"))
```

Industrial engineering

</div>
