
## Chapter 5
Returning to our goal of looking at state-level trends, let's add `Admin1Name` to our `group_by` function arguments. We'll also alter the name of the data frame to indicate that it contains state data.

```{R}
annual_state_counts <- read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue))
```

If you now run

```{r}
annual_state_counts
```

You'll see that we have three variable columns, `year`, `Admin1Name`, `count`. `group_by` took our data and grouped it by the `year` and `Admin1Name` variables. Let's use `geom_line` to look at the temporal trends of Lyme Disease over time for each state

```{r}
ggplot(annual_state_counts, aes(x=year, y=count, color=Admin1Name)) +
	geom_line()
```

Epic. Fail. It looks like a handful of states really exploded over time, but because there's so many colors and lines, it's impossible to figure out which shade of purple represents the state with the highest levels of Lyme Disease. Is that Pennsylvania? Rhode Island? South Dakota? I have my suspicions, but I'm not 100%. Returning to our motivation, we were blown away by the temporal trends depicted by a heatmap. Let's make a heatmap using the `geom_tile` function. As before, we can use `?geom_tile` to learn about the various aesthetics. The basics are `x`, `y`, and `fill`. Can you figure out what columns in `annual_state_counts` should correspond to each aesthetic? As before, we want `year` mapped to `x`. Now we're going to "flip" `y` and `fill` to represent `Admin1Name` and `count`, respectively

```{r}
ggplot(annual_state_counts, aes(x=year, y=Admin1Name, fill=count)) +
	geom_tile()
```

There's a lot to like about this way of viewing the data over the previous line plot. It's clear that Pennsylvania was the state that exploded with cases. But there's a lot to not like about this approach. If I try to find Michigan, it's hard (for me) to track the row for Michigan across without worrying that I'm looking at Massachusetts or Maryland. Because we suspect that the spread is driven, in part, by geography, we might want to group the states in a more interesting way. It's also hard to visually interpolate the color gradient to know how many cases there are in Michigan, Missouri, or Pennsylvania.

I'd like to return to a line plot to represent these data, but I only want to depict a handful of states. To do this, we'll need to learn more about the `filter` function. As the name suggests, this function filters the dataset to retain those rows that satisfy values of interest. We've already been using `filter` to retain those rows that are `PartOfCumulativeCountSeries`. This means that the value in `CountValue` column for that row are for the year rather than for a shorter time period (because states vary in how they report the data, I chose to go with these annual data). I can modify the existing `filter` function arguments or call the function again with different arguments. These both give me data for Michigan

```{r}
read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries, Admin1Name=="MICHIGAN")
```

```{r}
read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	filter(Admin1Name=="MICHIGAN")
```

Can you figure out how to know that these commands did what we hoped? If you were thinking about the `glimpse` or `View` functions, good work! An easier way would be to use the `count` function

```{r}
read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries, Admin1Name=="MICHIGAN") %>%
	count(Admin1Name)
```

In the first example we ask for those rows where `PartOfCumulativeCountSeries` is `TRUE` and where `Admin1Name` has the value of `"MICHIGAN"`. Alternatively stated, we are asking where `PartOfCumulativeCountSeries` is `TRUE` and where `Admin1Name=="MICHIGAN"` is `TRUE`. The comma in that first example can be spoken as "and" when you read the syntax. To make this more clear, I prefer to use the `&` symbol, which means "and".

```{r}
read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries & Admin1Name=="MICHIGAN") %>%
	count(Admin1Name)
```

We also often want to ask "or" type questions. For example, "Does this row contain data from Michigan, Missouri, or Pennsylvania?" To illustrate how to do this, we'll start by writing two `filter` functions and then show how we can consolidate them into one. Just as `&` represents "and", in R, `|` represents "or"

```{r}
read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	filter(Admin1Name=="PENNSYLVANIA" | Admin1Name=="MICHIGAN" | Admin1Name=="MISSOURI" ) %>%
	count(Admin1Name)
```

It is perfectly acceptable to have two filter lines as we do here. Perhaps you'd rather only have one. Keeping in mind [order of operations in math](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Syntax.html) does what's in parentheses before the `==`, we can use parentheses to bracket our query

```{r}
read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries & (Admin1Name=="PENNSYLVANIA" | Admin1Name=="MICHIGAN" | Admin1Name=="MISSOURI")) %>%
	count(Admin1Name)
```

There is a tradeoff between readability and code efficiency, but the key is that the syntax gives the right answer. Let's see about making this a line plot, by revisiting our code from the beginning of this chapter.

```{r}
annual_state_counts <- read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries & (Admin1Name=="PENNSYLVANIA" | Admin1Name=="MICHIGAN" | Admin1Name=="MISSOURI")) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue))

ggplot(annual_state_counts, aes(x=year, y=count, color=Admin1Name)) +
	geom_line() +
	scale_y_continuous(limits=c(0,NA)) +
	scale_x_continuous(breaks=c(1990, 1995, 2000, 2005, 2010, 2015)) +
	labs(x="Year",
			y="Annual number of cases",
			title="The number of Lyme disease cases continues to increase each year") +
	theme_classic()
```

Slick, eh? Michigan and Missouri clearly don't have anywhere the number of cases as Pennsylvania. Let's focus on Michigan and ask about the number of cases in the states that are around it - Ohio, Indiana, Illinios, and Wisconsin. Now we want to build filter the data to match one of five states. That's going to get a little long. Let's learn a new approach!

Remember that `filter` is looking for values that evaluate to `TRUE` or `FALSE` - these are "logicals". We can use an R function to test whether the values in a variable are in a set of values. This function, `%in%`, has a weird syntax

```{r}
favorite_letters <- c("p", "d", "s")

letters %in% favorite_letters
```

First, `letters` is a variable built into R that contains the 26 lowercase letters in the Roman alphabet (`LETTERS` has the uppercase version). We can read the last statement as "are the values in `letters` also in `favorite_letters`?" Let's apply this to our problem using two calls to `filter`

```{r}
upper_midwest_states <- c("MICHIGAN", "OHIO", "INDIANA", "WISCONSIN")

annual_state_counts <- read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
 	filter(Admin1Name %in% upper_midwest_states) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue))

ggplot(annual_state_counts, aes(x=year, y=count, color=Admin1Name)) +
	geom_line() +
	scale_y_continuous(limits=c(0,NA)) +
	scale_x_continuous(breaks=c(1990, 1995, 2000, 2005, 2010, 2015)) +
	labs(x="Year",
			y="Annual number of cases",
			title="The number of Lyme disease cases in Wisconsin has been higher\nthan other states in the upper midwest") +
	theme_classic()
```

Questions
* New England traditionally consists of Maine, Vermont, New Hampshire, Massachusetts, Rhode Island, and Connecticut. Lyme, Connecticut is the namesake of the disease. Can you modify our code to make a line plot for these states?

```{r}
new_england <- c("MAINE", "VERMONT", "NEW HAMPSHIRE", "MASSACHUSETTS", "RHODE ISLAND", "CONNECTICUT")

annual_state_counts <- read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
 	filter(Admin1Name %in% new_england) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue))

ggplot(annual_state_counts, aes(x=year, y=count, color=Admin1Name)) +
	geom_line() +
	scale_y_continuous(limits=c(0,NA)) +
	scale_x_continuous(breaks=c(1990, 1995, 2000, 2005, 2010, 2015)) +
	labs(x="Year",
			y="Annual number of cases",
			title="Connecticut continues to have a significant number of Lyme Disease cases") +
	theme_classic()
```

* 2016 was the last year that we have data for in this dataset. Can you make a dot plot with all of the possible `Admin1Name` values on the y-axis, the `count` values on the x-axis for 2016? Give the bars a color that you like.

```{r}
state_counts <- read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	filter(year == 2016) %>%
	group_by(Admin1Name) %>%
	summarize(count = max(CountValue))

ggplot(state_counts, aes(y=Admin1Name, x=count)) +
	geom_point() +
	scale_x_continuous(limits=c(0,NA)) +
	labs(x="Number of cases for 2016",
			y="",
			title="Pennsylvania has more cases than any other state") +
	theme_classic()
```


We're going to move on to a different dataset in the next section. Before we do, let's make a map of the data! Below is a code chunk to generate a map of the US with each state color coded by the number of cases reported in 2015.

```{r}
states_map <- map_data("state")

read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue)) %>%
	filter(year == 2010) %>%
	ggplot(aes(map_id = tolower(Admin1Name), fill=count)) +
		geom_map(map=states_map) +
		expand_limits(x = states_map$long, y = states_map$lat)
```

Take a moment and make a make a list of...
* Critiques of this map
* Questions this map raises for you
* Things in the code chunk you understand, have a good guess at, and are lost on

Here are some of my critiques...
* Most of the data for the US is uninteresting, could we tighten the limits to look at the the most interesting states?
* There is missing data that are probably important - where is New York state?
* Only one year's worth of data (2010)
* The color gradient leaves a bit to be desired

Here are some of my questions...
* The data are total counts, it would be better to have a per capita rate for each state
* What would a map look like for West Nile Virus?
* Could I generate an animation of the area east of Wisconsin and north of the Mason-Dixon line?

Things we've seen already and discussed...
* `library`, `<-`, `filter`, `group_by`, `summarize`, `%>%`, `ggplot`, `aes`, `fill`, `+`, `col_logical`

Things you can probably guess at...
* `map_data`, `tolower`, `geom_map`, `map`

Things that still might look foreign...
* `$`, `expand_limits`, `mutate`


Questions
* Can you use `?tolower` to confirm your intuition for what this function does? If you want a string to be in all caps, what function would you use? From the documentation for `tolower`, can you figure out how you might convert the sequence "ATGCCTTG" to "TACGGAAC"?

```{r}
quiet <- "hello, i don't like to yell"
toupper(quiet)

dna <- "ATGCCTTG"
chartr("ATGC", "TACG", dna)
```

* Most of the data for the US is uninteresting, could we tighten the limits to look at the the most interesting states? To do this we will modify the arguments for `expand_limits`. Instead of using `x = states_map$long, y = states_map$lat`, we will want to use something like `x = c(?, ?), y = c(?, ?)`. See if you can pick the best values to focus on those states from Wisconsin to Maine and from the Mason-Dixon line north.

```{r}
states_map <- map_data("state")

read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue)) %>%
	filter(year == 2010) %>%
	ggplot(aes(map_id = tolower(Admin1Name), fill=count)) +
		geom_map(map=states_map) +
		expand_limits(x = c(-93, -67), y = c(36.5,47.5))
```


* To make things even more interesting, could you make a map for your home state? You'll need to add an argument to `map_data` so be sure to use `?map_data` to help figure it out.

```{r}
states_map <- map_data("state", region = "michigan")

read_csv("project_tycho/US.23502006.csv",
					col_type=cols(PartOfCumulativeCountSeries = col_logical())) %>%
	filter(PartOfCumulativeCountSeries) %>%
	mutate(year = year(PeriodStartDate+7)) %>%
	group_by(year, Admin1Name) %>%
	summarize(count = max(CountValue)) %>%
	filter(year == 2010) %>%
	ggplot(aes(map_id = tolower(Admin1Name), fill=count)) +
		geom_map(map=states_map) +
		expand_limits(x = states_map$long, y = states_map$lat)
```
